---
title: 'Public health and economic consequnces of the severe weather events across the United States'
output:
  html_document:
    keep_md: yes
---

## Summary/Synopsis   
This report focuses on analysis of storms and other severe weather events that cause both public health and economic problems for communities and municipalities. We have analyzed the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database for severe events that result in fatalities, injuries, property and crop damages.

We show that tornadoes are the most harmful storm event for the population health (as measured by injuries and fatalities). Heat/excessive heat was the largest single cause of fatalities in 1995.  Flooding, thunderstorms and lightning  make notable contribution to the injury and fatality totals. Texas is number one state by number of fatalities and injuries.

Floods are the primary cause of the property damage, with tornados and hurricanes being close. Drought, flood and ice storms are the primary causes of the crop damage in the US.

## Data Processing   

Initializing and loading required packages for processing and analysis
```{r initiallise}
require(knitr, quietly=TRUE)
opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE)
```

```{r load libraries}
require(dplyr, quietly=TRUE)
require(lubridate, quietly=TRUE)
require(ggplot2, quietly=TRUE)
require(grid, quietly=TRUE)
require(gridExtra, quietly=TRUE)
require(scales, quietly=TRUE)
require(maps, quietly=TRUE)
data(state)
```

Downloading the databases file and load it into a data frame:
```{r dataload, results='hide', cache=TRUE}
# Load the data from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
fileUrl  <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
fileName <- "stormData.csv.bz2"

if(!file.exists(fileName)) {
    download.file(fileUrl, destfile = fileName, method = "curl")
}

stormDF <- read.csv(fileName, stringsAsFactors=FALSE)
```

#### Creating tidy data set   

The next few steps are designed to produce the data set containing mainly variables used in the data analysis grouped by state, year and storm / event type to identify events that result in most fatalities, injuries, property and crop damage across United States.   

Analyzing the structure of the Storm Data set
```{r data structure}
str(stormDF)
```

Tidying up the variable names and adding "year" column to enable easier analysis per year
```{r adding year}
names(stormDF) <- tolower(names(stormDF)) # tidy up the variable names
stormDF$evtype <- toupper(stormDF$evtype)
stormDF$year <- year(as.Date(stormDF$bgn_date, "%m/%d/%Y"))
```

##### Converting state abbreviation to the state names   
Storm DB contains data for all official US territories from 1950 to the end of 2011. We will specifically focus our analysis only on canonical 50 States.   

Comparing US State abbreviations in the data set with the canonical 50 States. 
```{r state names}
unique(stormDF$state); length(unique(stormDF$state))
state.abb; length(state.abb)
states <- data.frame(state = state.abb, 
                     state_name = tolower(state.name), 
                     stringsAsFactors = FALSE)
```

See references for more infomration on US official states and territories.

We will add a column with US official state names to enable per state analysis of the events.
```{r adding state names}
# NOTE: left_join hangs R session if state_name column already exist in stormDF data frame
stormDF$state_name <- NULL
stormDF <- left_join(stormDF, states, by="state")
```


##### Damage information analysis and adjustment   

We need to explore and normalize the exponents of the property and crop damages first before computing the damages. After we will need to adjust the damage cost using Consumer Price Index yearly index to account for inflation.   

First we notice discrete nature of the exponent data and some garbage that needs to be accounted for while computing the cost of property and crop damages.   

```{r damage}
unique(stormDF$propdmgexp)
unique(stormDF$cropdmgexp)
```

We will factorize the exponents and map them to the numeric values and compute property and crop damages.   

```{r damage exponent}
levels(as.factor(stormDF$propdmgexp))
levels(as.factor(stormDF$cropdmgexp))
```

Creating numerical exponents vectors that correspond to the above, converting property and crop damages into proper numerical form for further analysis. 

```{r exponent conversion}
numpropexp <- c(0, 0, 0, 0, 0, 1:8, 9, 2, 2, 3, 6, 6)
numcropexp <- c(0, 0, 0, 2, 9, 3, 3, 6, 6)

stormDF$propdamage <- stormDF$propdmg * 10^numpropexp[as.numeric(as.factor(stormDF$propdmgexp))]
stormDF$cropdamage <- stormDF$cropdmg * 10^numcropexp[as.numeric(as.factor(stormDF$cropdmgexp))]
```

To account for inflation over the years, we use Consumer Price Index to convert the damage cost to 2011 dollar index and calculate the damage costs for the entire Storm dataset to enable accurate year to year damage cost analysis.   


```{r}
# compute adjustment factor based on Consumer Price Index data (CPI)
cpi <- read.csv("http://research.stlouisfed.org/fred2/data/CPIAUCSL.csv", header = TRUE)
cpi$year <- year(cpi$DATE)
indx <- cpi %>% group_by(year) %>% summarize(indx = mean(VALUE))
indx$indx <- indx$indx[indx$year == max(stormDF$year)]/indx$indx

# adjsut costs to 2011 dollars
stormDF$propdamage <- stormDF$propdamage * left_join(stormDF, indx, by="year")$indx
stormDF$cropdamage <- stormDF$cropdamage * left_join(stormDF, indx, by="year")$indx
```

For our tidy data set we are selecting only the variables that are used in further analysis

```{r}
stormDF <- select(stormDF, 
                  state_name, year, evtype, 
                  fatalities, injuries, propdamage, cropdamage)
str(stormDF)
head(stormDF,20)
```


#### Storm Events categories normalization

We observe that number of unique severe weather event types recorded in the Storm Database is significantly larger than 96 described in the Storm Data Event Table from "NATIONAL WEATHER SERVICE INSTRUCTION 10-1605" (see References section)

```{r}
length(unique(stormDF$evtype))
```

Looking at the frequency of terms used and comparing event names to the documented standard 96 types we observe a number of duplicate types that need to be combined to accurately reflect the total impact form a given type of an event.

```{r}
stdf <- stormDF %>%
    group_by(evtype) %>%
    summarise(n = n()) %>%
    arrange(desc(n))
stdf
```

While categorizing all `898` recorded event types into `96` standard ones is far beyond this particular report, we are going to account for some duplication and consolidate events across most frequent ones.


```{r}
evtlookup <- function(s, v = T) {
    unique(grep(toupper(s), toupper(stormDF$evtype),value = v, perl=T))
}
```

```{r cache=TRUE}
stdf[1:50,]
# regex bulk updates
stormDF$evtype[evtlookup("heavy rain", v=F)] = "HEAVY RAIN"
stormDF$evtype[evtlookup("hail", v=F)] = "HAIL"
stormDF$evtype[evtlookup("tornado", v=F)] = "TORNADO"

stormDF$evtype[evtlookup("drought", v=F)] = "DROUGHT"
stormDF$evtype[evtlookup("(record|excessive|extreme).*heat", v=F)] = "EXCESSIVE HEAT"
stormDF$evtype[evtlookup("heat wave", v=F)] = "HEAT"

stormDF$evtype[evtlookup("flash", v=F)] = "FLASH FLOOD"
stormDF$evtype[evtlookup("coastal flood", v=F)] = "COASTAL FLOOD"
stormDF$evtype[evtlookup("^(?!flash|coastal).*flood", v=F)] = "FLOOD"

stormDF$evtype[evtlookup("ice storm", v=F)] = "ICE STORM"
stormDF$evtype[evtlookup("avalanche", v=F)] = "AVALANCHE"
stormDF$evtype[evtlookup("blizzard", v=F)] = "BLIZZARD"
stormDF$evtype[evtlookup("(extreme).*wind.*ch", v=F)] = "EXTREME COLD/WIND CHILL"
stormDF$evtype[evtlookup("(extreme|record|excessive).*cold", v=F)] = "EXTREME COLD/WIND CHILL"
stormDF$evtype[evtlookup("^(?!extreme).*wind ch", v=F)] = "COLD/WIND CHILL"
stormDF$evtype[evtlookup("(heavy|record|excessive).*snow", v=F)] = "HEAVY SNOW"
stormDF$evtype[stormDF$evtype == "SNOW"] = "HEAVY SNOW"
stormDF$evtype[evtlookup("freeze|frost", v=F)] = "FROST/FREEZE"
stormDF$evtype[evtlookup("sleet", v=F)] = "SLEET"
stormDF$evtype[evtlookup("winter weather", v=F)] = "WINTER WEATHER"

stormDF$evtype[evtlookup("hurricane|typhoon", v=F)] = "HURRICANE/TYPHOON"
stormDF$evtype[evtlookup("^(?!marine).*thunderstorm", v=F)] = "THUNDERSTORM WIND"
stormDF$evtype[evtlookup("^(?!marine).*tstm", v=F)] = "THUNDERSTORM WIND"
stormDF$evtype[stormDF$evtype == "MARINE TSTM WIND"] = "MARINE THUNDERSTORM WIND"
stormDF$evtype[evtlookup("^(high|strong).*wind", v=F)] = "STRONG WIND"
stormDF$evtype[stormDF$evtype == "WIND"] = "STRONG WIND"
stormDF$evtype[stormDF$evtype == "STORM SURGE"] = "STORM SURGE/TIDE"

stormDF$evtype[evtlookup("(dense).*fog", v=F)] = "DENSE FOG"
stormDF$evtype[evtlookup("^(?!dense).*fog", v=F)] = "FREEZING FOG"
stormDF$evtype[evtlookup("fire", v=F)] ="WILDFIRE"
stormDF$evtype[evtlookup("current", v=F)] ="RIP CURRENT"

```

#### Finalizing the tidy data set

Lastly we will group events by year and state and limit our analysis only to 50 canonical states including only events that resulted in one or more health and/or economic problems(fatalities, injuries, property and crop damage)

```{r}
stormDF <- stormDF %>% 
    filter(!(is.na(stormDF$state_name))) %>%
    filter(fatalities > 0 |
             injuries > 0 |
           propdamage > 0 |
           cropdamage > 0) %>%
    group_by(year, state_name, evtype) %>%
    summarize(fatalities = sum(fatalities),
                injuries = sum(injuries),
              propdamage = sum(propdamage),
              cropdamage = sum(cropdamage))
stormDF
```

## Results

#### Analysis of the number of fatalities and injuries

To aid a graphical representation of the results we will compute a few summaries of the number of fatalities and injuries for event totals across US, totals breakdown by state and events breakdown by year.


```{r}
fatal_usa <- stormDF %>% 
    group_by(evtype) %>%
    summarise(fatal_total = sum(fatalities)) %>%
    arrange(desc(fatal_total)) %>%
    mutate(rank=1:nrow(.))
```

```{r}
injury_usa <- stormDF %>%
    group_by(evtype) %>%
    summarise(injury_total = sum(injuries)) %>%
    arrange(desc(injury_total)) %>%
    mutate(rank=1:nrow(.))
```

```{r}
fatal_state <- stormDF %>% 
    group_by(state_name) %>% 
    summarize(fatal_total = sum(fatalities)) %>%
    arrange(desc(fatal_total))
```

```{r}
injury_state <- stormDF %>% 
    group_by(state_name) %>%
    summarise(injury_total = sum(injuries)) %>%
    arrange(desc(injury_total))
```

```{r}
fatal_year <- stormDF %>%
    filter(evtype %in% fatal_usa$evtype[1:5]) %>%
    group_by(year, evtype) %>%
    summarise(fatal_total=sum(fatalities)) %>%
    arrange(year)
```

```{r}
injury_year <- stormDF %>%
    filter(evtype %in% injury_usa$evtype[1:5]) %>%
    group_by(year, evtype) %>%
    summarise(injury_total=sum(injuries)) %>%
    arrange(year)
```

```{r}
states_map <- map_data("state")

theme_map <- theme(plot.title = element_text(size = 12),
                   axis.title = element_blank(),
                   axis.text = element_blank(),
                   axis.ticks = element_blank(),
                   panel.border = element_blank(),
                   panel.grid = element_blank(),
                   panel.background = element_rect(fill = "white",  colour = NA))

theme_bar <- theme(axis.title.y = element_blank(),
                   axis.ticks.y = element_blank(),
                   panel.grid.major.y = element_blank(),
                   panel.border = element_blank()
                   )
```


The following figure summarizes the analysis of the health consequences due to severe weather. It is organized by state, by event type, by year and event type  showing total number of fatalities and injuries.   

```{r fig.align='center', fig.width=10, fig.height=12, fig.cap="Health impact of the sever weather events"}
g1 <- ggplot(fatal_state, aes(map_id = state_name)) + 
    geom_map(aes(fill = fatal_total), map = states_map) +
    scale_fill_gradient("Total", low = "#EEEEEE", high = "#990000") +
    expand_limits(x = states_map$long, y = states_map$lat) +
    ggtitle("Fatalities per state across USA") +
    theme_map
   
g2 <- ggplot(injury_state, aes(map_id = state_name)) + 
    geom_map(aes(fill = injury_total), map = states_map) +
    scale_fill_gradient("Total", low = "#EEEEEE", high = "#003366") +
    expand_limits(x = states_map$long, y = states_map$lat) +
    ggtitle("Injuries per state across USA") +
    theme_map

g3 <- ggplot(fatal_usa[1:20,], aes(x=rank, y=fatal_total, fill=fatal_total)) +
    geom_bar(stat="identity") + coord_flip() +
    scale_fill_gradient("Count", low = "pink", high = "#990000") +
    scale_x_reverse(breaks=1:20, labels=fatal_usa$evtype[1:20]) + 
    scale_y_continuous(breaks=c(1000, 2000, 3000, 4000, 5000)) +
    ggtitle("Top 20 most fatal even types") +
    ylab("Total number of fatalities recorded") +
    theme_bw(base_size = 10) +
    theme_bar
    
g4 <- ggplot(injury_usa[1:20,], aes(x=rank, y=injury_total, fill=injury_total)) +
    geom_bar(stat="identity") + coord_flip() +
    scale_fill_gradient("Count", low = "#66CCFF", high = "#003366") +
    scale_x_reverse(breaks=1:20, labels=injury_usa$evtype[1:20]) + 
    scale_y_continuous(breaks=c(20000, 40000, 60000, 80000)) +
    ggtitle("Top 20 event most harmful event types") +
    ylab("Total number of injuries recorded") +
    theme_bw(base_size = 10) +
    theme_bar

g5 <- ggplot(fatal_year, aes(x=year, y=fatal_total, colour = evtype)) +
    geom_line(size = 2, alpha = 0.8) +
    scale_colour_discrete(name="Event Type", breaks=fatal_usa$evtype[1:5]) +
    scale_x_continuous(breaks=c(1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
    ggtitle("Fatalities for top 5 types") +
    xlab("Year") + ylab("Fatalities") +
    theme_bw(base_size = 10) +
    theme(legend.position="left")

g6<- ggplot(injury_year, aes(x=year, y=injury_total, colour = evtype)) +
    geom_line(size = 2, alpha = 0.8) +
    scale_colour_discrete(name="Event Type", breaks=injury_usa$evtype[1:5]) +
    scale_x_continuous(breaks=c(1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
    ggtitle("Injuries for top 5 types") +
    xlab("Year") + ylab("Injuries") +
    theme_bw(base_size = 10) +
    theme(legend.position="left")
#g6
grid.arrange(g1, g2, g3, g4, g5, g6, ncol=2)
```

##### Discussion

The figure above shows that tornadoes are the most harmful storm event for the population health (as measured by injuries and fatalities). Heat/excessive heat was the largest single cause of fatalities in 1995.  Flooding, thunderstorms and lightning  make notable contribution to the injury and fatality totals. Texas and Illinois clearly stand out as the states with the largest numbers of fatalities.  Texas is also number one state in number of injuries caused by severe weather events.  

```{r}
fatal_year$evtype[which.max(fatal_year$fatal_total)]
fatal_year$year[which.max(fatal_year$fatal_total)]
injury_year$evtype[which.max(injury_year$injury_total)]
injury_year$year[which.max(injury_year$injury_total)]
```

As shown, heat in 1995 was the primary cause of the maximum fatalities within one single year. The maximum number of injuries within a single year is caused by 1974 tornados.   


#### Analysis of the property and crop damage   

To aid a graphical representation of the results we will compute a few summaries of the property and crop damages for event totals across US, totals breakdown by state and events breakdown by year.   

```{r}
prop_usa <- stormDF %>% 
    group_by(evtype) %>%
    summarise(prop_total = sum(propdamage)) %>%
    arrange(desc(prop_total)) %>%
    mutate(rank=1:nrow(.))
```

```{r}
crop_usa <- stormDF %>%
    group_by(evtype) %>%
    summarise(crop_total = sum(cropdamage)) %>%
    arrange(desc(crop_total)) %>%
    mutate(rank=1:nrow(.))
```

```{r}
prop_state <- stormDF %>% 
    group_by(state_name) %>% 
    summarize(prop_total = sum(propdamage)) %>%
    arrange(desc(prop_total))
```

```{r}
crop_state <- stormDF %>% 
    group_by(state_name) %>%
    summarise(crop_total = sum(cropdamage)) %>%
    arrange(desc(crop_total))
```

```{r}
prop_year <- stormDF %>%
    filter(evtype %in% prop_usa$evtype[1:5]) %>%
    group_by(year, evtype) %>%
    summarise(damage=sum(propdamage)) %>%
    arrange(year)
```

```{r}
crop_year <- stormDF %>%
    filter(evtype %in% crop_usa$evtype[1:5]) %>%
    group_by(year, evtype) %>%
    summarise(damage=sum(cropdamage)) %>%
    arrange(year)
```

The following figure summarizes the analysis of the economic consequences due to severe weather. It shows by state, by event type, by year and event type total cost of adjusted damages to property and crop.

```{r fig.align='center', fig.width=10, fig.height=12, fig.cap="Economic impact of the sever weather events"}
g7 <- ggplot(prop_state, aes(map_id = state_name)) + 
    geom_map(aes(fill = prop_total/10^9), map = states_map) +
    scale_fill_gradient("Cost, B", low = "#EEEEEE", high = "#660066") +
    expand_limits(x = states_map$long, y = states_map$lat) +
    ggtitle("Property damage per state across USA") +
    theme_map
   
g8 <- ggplot(crop_state, aes(map_id = state_name)) + 
    geom_map(aes(fill = crop_total/10^9), map = states_map) +
    scale_fill_gradient("Cost, B", low = "#EEEEEE", high = "#FF3300") +
    expand_limits(x = states_map$long, y = states_map$lat) +
    ggtitle("Crop damage per state across USA") +
    theme_map

g9 <- ggplot(prop_usa[1:20,], aes(x=rank, y=prop_total, fill=prop_total/10^9)) +
    geom_bar(stat="identity") + coord_flip() +
    scale_fill_gradient("Cost, B", low = "pink", high = "#660066") +
    scale_x_reverse(breaks=1:20, labels=prop_usa$evtype[1:20]) + 
    scale_y_continuous(labels=c(0, 50, 100, 150, 200)) +
    ggtitle("Top 20 property damage event types") +
    ylab("Total cost of property damange, Billions") +
    theme_bw(base_size = 10) +
    theme_bar
    
g10 <- ggplot(crop_usa[1:20,], aes(x=rank, y=crop_total, fill=crop_total/10^9)) +
    geom_bar(stat="identity") + coord_flip() +
    scale_fill_gradient("Cost, B", low = "#FFFF00", high = "#FF3300") +
    scale_x_reverse(breaks=1:20, labels=crop_usa$evtype[1:20]) + 
    scale_y_continuous(labels=c(0, 5, 10, 15, 20)) +
    ggtitle("Top 20 crop damage event types") +
    ylab("Total cost of crop damage, Billions") +
    theme_bw(base_size = 10) +
    theme_bar

g11 <- ggplot(prop_year, aes(x=year, y=damage, colour = evtype)) +
    geom_line(size = 2, alpha = 0.8) +
    scale_colour_discrete(name="Event Type", breaks=prop_usa$evtype[1:5]) +
    scale_x_continuous(breaks=c(1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
    scale_y_continuous(labels=c(0, 50, 100, 150)) +
    ggtitle("Property damage for top 5 types") +
    xlab("Year") + ylab("Property damage, Billions") +
    theme_bw(base_size = 10) +
    theme(legend.position="left")

g12<- ggplot(crop_year, aes(x=year, y=damage, colour = evtype)) +
    geom_line(size = 2, alpha = 0.8) +
    scale_colour_discrete(name="Event Type", breaks=crop_usa$evtype[1:5]) +
    scale_x_continuous(limits=c(1950, 2010), breaks=c(1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
    scale_y_continuous(labels=c(0, 2, 4, 6, 8)) +
    ggtitle("Crop damage for top 5 types") +
    xlab("Year") + ylab("Crop damages, Billions") +
    theme_bw(base_size = 10) +
    theme(legend.position="left")

grid.arrange(g7, g8, g9, g10, g11, g12, ncol=2)
```

##### Discussion  

The figure above demonstrates that floods are the primary cause of the property damage, with tornados and hurricanes being second close. Drought, flood and ice storms are the primary causes of the crop damage in the US. California clearly stands out as the state with most property damage overall. Texas, Mississippi and Illinois have most crop damage overall.   

```{r}
prop_year$evtype[which.max(prop_year$damage)]
prop_year$year[which.max(prop_year$damage)]
crop_year$evtype[which.max(crop_year$damage)]
crop_year$year[which.max(crop_year$damage)]
```

As shown, maximum property damage within one single year is caused by flood in 2006. The maximum crop damage within a single year is caused by 1993 flood.   

## References

http://en.wikipedia.org/wiki/List_of_U.S._state_abbreviations

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf